Is this normal? A new projection pursuit index to assess a sample against a multivariate null distribution

### 2025-01-03 paper.qmd
copying edits from earlier

### Projecting an ellipsoid

Let $x$ be a \pD{} vector. A \pD{} ellipsoid corresponding to a given variance-covariance ($\Sigma$) is defined by

@@

where $c$ a constant that depends on a specific confidence level.

\begin{theorem}
  The projection of this \pD{} ellipsoid in \gD{} has the equation
\begin{equation}
(\mathbfit{y} - \mathbfit{\mu}_p)(P^T \Sigma P)^{-1}(\mathbfit{y} - \mathbfit{\mu}_p)^T = c^2.
\end{equation}
\end{theorem}
\begin{proof}
The projection of an ellipsoid onto \gD{} is an ellipse, where the curve of the ellipse is defined through the set of points $\mathbfit{x}$ for which the gradient is parallel to the projection plane. That is, the curve 
consists of $\mathbfit{x}$ satisfying

\begin{equation}\nabla (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T = 2 (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1} = 2\mathbfit{s}P^T
\end{equation}

for some $\mathbfit{s} \in \mathbb{R}^2$, where $P$ is a $(p\times 2)$ orthonormal basis defining the
projection. We can write $\mathbfit{x}-\mathbfit{\mu} = \mathbfit{s} P^T \Sigma$.
Making this substitution in Equation~(\ref{eq:pd}) yields

@@

This means the matrix $(P^T \Sigma P)^{-1}$ is defining the ellipse in the \gD{} projection. In general $c$ could be any constant, but typically we would select it as a quantile of the $\chi^2$ distribution, so that the size of the ellipse corresponds to a selected probability.

>>>>

### Projecting an ellipsoid
Let \(\mathbfit{x}\) be a \(p\) dimensional vector. A \pD{} ellipsoid corresponding to a given
variance-covariance (\(\Sigma\)) is described by the equation

@@

where \(c\) is a constant that depends on a specific confidence level.

Let \(P\) be a \((p\times 2)\) matrix, whose columns form an orthonormal basis for the \gD{} projection space. Let \(\mathbfit{y}\) be a 2 dimensional vector, and let \(\mathbfit{\mu}_p := \mathbfit{\mu} P\) denote the projected mean.
\begin{theorem}
  The projection of the \pD{} ellipsoid decribed in Equation~\ref{eq:pd} onto the \gD{} projection space described by \(P\) has the equation
\begin{equation}
(\mathbfit{y} - \mathbfit{\mu}_p)(P^T \Sigma P)^{-1}(\mathbfit{y} - \mathbfit{\mu}_p)^T = c^2.
\end{equation}
\end{theorem}
\begin{proof}
The projection of a hyperellipsoid onto \gD{} space is an ellipse. The boundary of this \gD{} ellipse is the orthogonal projection of select points on the boundary of the hyperellipsoid. For \(\mathbfit{x}\) on the hyperellipsoid, its orthogonal projection \(\mathbfit{x}P\) is on the boundary of the projected ellipse if and only if the tangent plane to the hyperellipsoid at \(\mathbfit{x}\) is perpendicular to the projection plane described by \(P\). As the hyperellipsoid is a level curve, described by Equation~(\ref{eq:pd}), this means the gradient of the left hand side of Equation~(\ref{eq:pd}) at \(\mathbfit{x}\) is parallel to the projection plane. Note that any vector parallel to the projection plane can be expressed as \(2\mathbfit{s}P^T\) for some \(\mathbfit{s} \in \mathbb{R}^2\).

Therefore, the boundary of the ellipse consists of the points \(\mathbfit{x}P\) for which $\mathbfit{x}$ satisfies

\begin{equation}
2\mathbfit{s}P^T = \nabla \left[(\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}(\mathbfit{x}-\mathbfit{\mu})^T\right] = 2 (\mathbfit{x}-\mathbfit{\mu}) \Sigma^{-1}
\label{eq:grad}
\end{equation}

for some \(\mathbfit{s} \in \mathbb{R}^2\). Rearranging this equation gives \(\mathbfit{x}-\mathbfit{\mu} = \mathbfit{s} P^T \Sigma\).
Making this substitution in Equation~(\ref{eq:pd}) yields

@@

This means the matrix \((P^T \Sigma P)^{-1}\) defines the ellipse in
the \gD{} projection. In general \(c\) could be any constant, but
typically we would select it as a quantile of the \(\chi^2\)
distribution, so that the size of the ellipse corresponds to a selected
probability.
